---
title: "BAM Thesis - timeseries"
author: "Lian van Nee"
date: "2025-04-28"
output: word_document
---

# Libraries

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(tsibble)
library(stringr)
library(readr)
library(purrr)
library(dplyr)
library(tidyr)
library(readxl)
library(forecast)
library(tseries)
library(strucchange)
library(ppcor)
library(conflicted)
library(zoo)
library(prophet)
library(Metrics)
library(ggplot2)
library(Matrix)
library(xgboost)
library(broom)


options(contrasts = c("contr.treatment", "contr.treatment"))
conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
```

# Importing & Cleaning Data

```{r}
"Importing Glims data"
# Data Loading I have split it up in smaller parts so that it can be uploaded to GIT
# List all the chunks
chunk_files <- list.files(
  path = here("Data", "Order_data"),
  pattern = "data_chunk_.*\\.csv",
  full.names = TRUE
)
# Load and combine the chunks back into one dataset
data <- do.call(rbind, lapply(chunk_files, read.csv))
# Remove empty rows
data <- data[rowSums(is.na(data)) != ncol(data), ]

# We start with 4577229 observations

"Making adjustments to data"
# Changing dates to data format
data$datum_creatie <- as.Date(data$datum_creatie, format = "%d/%m/%Y")
data$datum_afname <- as.Date(data$datum_afname, format = "%d/%m/%Y")

# Changing Cito to binary variable
data <- data %>%
  mutate(Cito = if_else(Cito_JN == "Cito", 1, 0)) %>% 
  select(-Cito_JN, -peildatum)

"Adding variables"

# Adding testing variable
data <- data %>%
  mutate(self_testing = ifelse(str_detect(afname_locatie, "^[0-9]"), 1, 0))

"Deleting huisbezoek, selftesting and Cito data"
data_clean <- data %>%
  filter((data$huisbezoek != 1 | is.na(data$huisbezoek)) &
  (data$self_testing != 1 | is.na(data$self_testing)) &
  (data$Cito != 1 | is.na(data$Cito)))
# nrow(data_clean) = 3782446
```

## Kennemerland Specific Data

```{r}
# Getting all Kennemerland Postal Codes
kennemerland_postalcodes <- data %>% 
  filter(BD_regio %in% c("Kennemerland Zuid", "Kennemerland Noord", "Hoofddorp")) %>% 
  distinct(postcode_patient) %>% 
  mutate(PC4 = as.character(postcode_patient)) %>%
  select(-postcode_patient)

# creating data_clean for Kennemerland
ken_data_clean <- data_clean %>% 
  filter(postcode_patient %in% kennemerland_postalcodes$PC4) %>%
  filter(!datum_creatie=="2025-04-25") # Delete 2025-04-25, as this day was not fully recorded
nrow(ken_data_clean)
# should be 1506410

write.csv(ken_data_clean, file = here::here("Data", "ken_order_data_clean.csv"), row.names = FALSE)
```

## Creating ken_daily orders

```{r}
ken_daily_orders <- ken_data_clean %>%
  group_by(datum_creatie) %>%
  summarise(order_count = n(), .groups = "drop")

ken_daily_orders %>%
  summary(order_count)

# Total number of days
nrow(ken_daily_orders)

# Check for missing days
all_dates <- data.frame(
  datum_creatie = seq.Date(from = as.Date("2022-01-01"),
                           to = as.Date("2025-04-24"),
                           by = "day"))

# Join with your existing dataset to find missing dates
missing_dates <- all_dates %>%
  anti_join(ken_daily_orders, by = "datum_creatie")

# View results
print(missing_dates) # THere are no missing dates
```

## Demand Distribution + Var per daytype

```{r summary statistics per weekday}
ken_daily_orders <- ken_data_clean %>%
  group_by(datum_creatie) %>%
  summarise(order_count = n(), .groups = "drop") %>%
  arrange(datum_creatie) %>%
  mutate(weekday = wday(datum_creatie, label = TRUE, abbr = FALSE))

# Summary statistics per weekday
ken_summary_table_per_day <- ken_daily_orders %>%
  group_by(weekday) %>%
  summarise(
    count = n(),
    mean = mean(order_count),
    sd = sd(order_count),
    min = min(order_count),
    q25 = quantile(order_count, 0.25),
    median = median(order_count),
    q75 = quantile(order_count, 0.75),
    max = max(order_count)
  ) %>%
  arrange(factor(weekday, levels = c(
    "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"
  )))

write.csv(ken_summary_table_per_day, here("summary_tables", "ken_summary_table_per_day.csv"), row.names = FALSE)
```

```{r histogram distribution of daily order count}
"Difference Weekend and Weekday"
ken_daily_orders <- ken_daily_orders %>%
  mutate(day_type = if_else(wday(datum_creatie, label = TRUE) %in% c("Sat", "Sun"),
                            "Weekend", "Weekday"),
         day_type = factor(day_type, levels = c("Weekend", "Weekday")))

ggplot(ken_daily_orders, aes(x = order_count, fill = day_type)) +
  geom_histogram(binwidth = 15, position = "identity", alpha = 0.9) +
  scale_fill_manual(values = c("Weekday" = "cornflowerblue", "Weekend" = "indianred")) +
  labs(title = "Distribution of Daily Order Counts by Day Type",
       x = "Order Count (per day)", y = "Number of Days",
       fill = "Day Type") +
  theme_minimal()

ggsave(
  filename = here("figures", "distribution_daily_order_counts_per_day_type.png"),
  width = 10, height = 6, dpi = 300
)
```

```{r}
# Add log-transformed order count
ken_daily_orders <- ken_daily_orders %>%
  mutate(log_order_count = log(order_count))

# Plot
ggplot(ken_daily_orders, aes(x = log_order_count, fill = day_type)) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.9) +
  scale_fill_manual(values = c("Weekday" = "cornflowerblue", "Weekend" = "indianred")) +
  labs(title = "Log-Transformed Distribution of Daily Order Counts by Day Type",
       x = "Log(Order Count)", y = "Number of Days",
       fill = "Day Type") +
  theme_minimal()

# Saving
ggsave(
  filename = here("figures", "distribution_daily_log_order_counts_per_day_type.png"),
  width = 10, height = 6, dpi = 300
)
```

```{r combining the two graphs into one}
library(patchwork)

ken_daily_orders <- ken_daily_orders %>%
  mutate(day_type = if_else(wday(datum_creatie, label = TRUE) %in% c("Sat", "Sun"),
                            "Weekend", "Weekday"),
         day_type = factor(day_type, levels = c("Weekend", "Weekday")),
         log_order_count = log(order_count))

# Plot 1: Original order count distribution
p1 <- ggplot(ken_daily_orders, aes(x = order_count, fill = day_type)) +
  geom_histogram(binwidth = 15, position = "identity", alpha = 0.9) +
  scale_fill_manual(values = c("Weekday" = "cornflowerblue", "Weekend" = "indianred")) +
  labs(title = "Raw Order Count", x = "Order Count (per day)", y = "Number of Days", fill = "Day Type") +
  theme_minimal()

# Plot 2: Log-transformed order count distribution
p2 <- ggplot(ken_daily_orders, aes(x = log_order_count, fill = day_type)) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.9) +
  scale_fill_manual(values = c("Weekday" = "cornflowerblue", "Weekend" = "indianred")) +
  labs(title = "Log-Transformed Order Count", x = "Log(Order Count)", y = "Number of Days", fill = "Day Type") +
  theme_minimal()

# Combine with patchwork
combined_plot <- p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

# Save using here()
ggsave(
  filename = here("figures", "combined_order_count_distribution.png"),
  plot = combined_plot,
  width = 10,  # adjust as needed for your thesis format
  height = 5,  # adjust as needed
  dpi = 300
)
```

## Interweekly patterns

```{r performing-tests}
"Adding DoW as variable"
ken_daily_orders <- ken_daily_orders %>% 
  mutate(
    day_of_week = as.factor(wday(datum_creatie, label = TRUE, week_start = 1))
  )

"Showing means per weekday"
ken_daily_orders %>%
  group_by(day_of_week) %>%
  summarise(
    mean_orders = mean(order_count),
    sd_orders = sd(order_count),
    .groups = "drop"
  )

"Performing an Anova test to see differences between weekdays"
weekdays_only <- ken_daily_orders %>%
  filter(day_of_week %in% c("Mon", "Tue", "Wed", "Thu", "Fri")) %>%
  mutate(day_of_week = factor(day_of_week)) %>%
  droplevels()

anova_weekdays <- aov(order_count ~ day_of_week, data = ken_daily_orders)
summary(anova_weekdays)
# Saving result so it can be inserted into text
anova_tidy <- tidy(anova_weekdays)
write.csv(anova_tidy, here::here("summary_tables", "anova_weekday.csv"), row.names = FALSE)

"Tukey Test"
tukey_result <- TukeyHSD(anova_weekdays)

# For thesis insertion
tukey_tidy <- broom::tidy(tukey_result)
write.csv(tukey_tidy, here::here("summary_tables", "tukey_weekday.csv"), row.names = FALSE)

```

```{r visualizing-week-patterns}
"Visualizing"
ken_daily_orders <- ken_data_clean %>%
  group_by(datum_creatie) %>%
  summarise(order_count = n()) %>%
  arrange(datum_creatie) %>%
  mutate(
    day_of_week = wday(datum_creatie, label = TRUE, abbr = FALSE, week_start = 1),
    week_start_date = floor_date(datum_creatie, unit = "week", week_start = 1),
    week_numeric = as.numeric(week_start_date),
    week_label = paste0(year(week_start_date), " W", sprintf("%02d", isoweek(week_start_date)))
  )

# Choosing colors
soft_colors <- c(
  "#3B82F6",  # vivid blue
  "#60A5FA",  # sky blue
  "#EC4899",  # pinkish red
  "#F87171",  # warm red
  "#FCD34D",  # soft yellow (adds contrast)
  "#6EE7B7",  # light green
  "#10B981"   # bold green
)

# Choose breaks and labels for legend
week_breaks <- sort(unique(ken_daily_orders$week_numeric))[seq(1, length(unique(ken_daily_orders$week_numeric)), length.out = 6)]
week_labels <- ken_daily_orders %>%
  filter(week_numeric %in% week_breaks) %>%
  distinct(week_numeric, week_label) %>%
  arrange(week_numeric) %>%
  pull(week_label)

# Plot
ggplot(ken_daily_orders, aes(x = day_of_week, y = order_count, group = week_start_date, color = week_numeric)) +
  geom_line(alpha = 0.8, size = 0.3) +
  scale_color_gradientn(
    colours = soft_colors,
    breaks = week_breaks,
    labels = week_labels,
    name = "Week"
  ) +
  labs(
    x = "Day of the Week",
    y = "Number of Orders",
    title = "Weekly Order Patterns by Day"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

ggsave(here("figures", "weekly_orders_plot.png"),width = 10, height = 6, dpi = 300)
```

## Monthly patterns

```{r analysis-monthly-patterns}
ken_daily_orders <- ken_daily_orders %>%
  mutate(
    log_order_count = log1p(order_count)
  )

"Adding Month as variable"
ken_daily_orders <- ken_daily_orders %>% 
  mutate(
    month = factor(month(datum_creatie, label = TRUE), levels = month.abb),
    year = year(datum_creatie)
  )

"Showing means and deviations per month"
monthly_summary <- ken_daily_orders %>%
  group_by(month) %>%
  summarise(
    mean_orders = mean(order_count),
    sd_orders = sd(order_count),
    .groups = "drop"
  )

print(monthly_summary)

"ANOVA test"
anova_month <- aov(order_count ~ month, data = ken_daily_orders)
summary(anova_month)

# Saving result so it can be inserted into text
anova_tidy <- tidy(anova_month)
write.csv(anova_tidy, here::here("summary_tables", "anova_month.csv"), row.names = FALSE)
```

```{r visualization-monhtly-patterns}
"Visualization"
monthly_total_year <- ken_daily_orders %>%
  mutate(month = factor(month(datum_creatie, label = TRUE), levels = month.abb)) %>%
  group_by(year, month) %>%
  summarise(total_orders = sum(order_count), .groups = "drop")

# Define your custom color palette
custom_colors <- c(
  "2022" = "#60A5FA",  # light blue
  "2023" = "#EC4899",  # pink
  "2024" = "#F87171",  # red
  "2025" = "#FCD34D"   # yellow
)

# Plot
ggplot(monthly_total_year, aes(x = month, y = total_orders, group = factor(year), color = factor(year))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors, name = "Year") +
  labs(
    title = "Total Monthly Order Volume by Year",
    x = "Month",
    y = "Total Orders per Month"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

ggsave(here("figures", "monthly_total_order_volume_plot.png"),width = 10, height = 6, dpi = 300)
```

## Autocorrelation Analysis

```{r}
library(forecast)
library(patchwork)

p1 <- ggAcf(ken_daily_orders$order_count) + ggtitle("ACF of Daily Order Count")
p2 <- ggPacf(ken_daily_orders$order_count) + ggtitle("PACF of Daily Order Count")

p1 / p2
combined_plot <- p1 / p2

# Save
ggsave(
  filename = here::here("figures", "acf_pacf_combined.png"),
  plot = combined_plot,
  width = 8, height = 6
)
```

# External Variables

Due to the large diference in order volumes between weekdays (\~1800 orders) and weekends (\~50 orders), the variance in the response variable is expected to be non-constant. To stabilize the variance and reduce skewness, a log transformation of the response variable was applied prior to modeling. This means that also in exploratory research, the linear models looking at the usefulness of an external factor will also be conducted using log orders.

```{r creating log orders}
ken_daily_orders <- ken_daily_orders %>%
  mutate(log_orders = log(order_count + 1))
```

## Public & School holidays

```{r importing data and adding to df}
"import datasets"
public_holidays_nl <- read_excel(here("Data", "External_Variables", "public_holidays_nl.xlsx"))
school_holidays <- read_excel(here("Data", "External_Variables", "dates_school_holidays.xlsx"))

public_holidays_nl$date <- as.Date(public_holidays_nl$date)
school_holidays$date <- as.Date(school_holidays$date)

"Adding variables to dataframe"
ken_daily_orders <- ken_daily_orders %>% 
  mutate(
    is_PublicHoliday = ifelse(datum_creatie %in% public_holidays_nl$date, 1, 0),
    is_SchoolHoliday = ifelse(datum_creatie %in% school_holidays$date, 1, 0),
  )
```

```{r regression}
"Correlation"
cor(ken_daily_orders %>% select(order_count, is_PublicHoliday), use = "complete.obs")
cor(ken_daily_orders %>% select(log_orders, is_PublicHoliday), use = "complete.obs")

"See if school/public holidays significantly improves the model"
model_1 <- lm(order_count ~ day_of_week, data = ken_daily_orders)

model_3 <- lm(order_count ~ day_of_week + is_SchoolHoliday, data = ken_daily_orders)
model_4 <- lm(order_count ~ day_of_week*is_SchoolHoliday, data = ken_daily_orders)
model_5 <- lm(order_count ~ day_of_week*is_SchoolHoliday + is_PublicHoliday, data = ken_daily_orders)

model_6 <- lm(order_count ~ day_of_week + is_PublicHoliday, data = ken_daily_orders)
model_7 <- lm(order_count ~ day_of_week*is_PublicHoliday, data = ken_daily_orders)
model_8 <- lm(order_count ~ day_of_week*is_PublicHoliday + is_SchoolHoliday, data = ken_daily_orders)

model_9 <- lm(order_count ~ day_of_week * (is_PublicHoliday + is_SchoolHoliday), data = ken_daily_orders)


"Log variation of model"
model_1 <- lm(log_orders ~ day_of_week, data = ken_daily_orders)

model_3 <- lm(log_orders ~ day_of_week + is_SchoolHoliday, data = ken_daily_orders)
model_4 <- lm(log_orders ~ day_of_week*is_SchoolHoliday, data = ken_daily_orders)
model_5 <- lm(log_orders ~ day_of_week*is_SchoolHoliday + is_PublicHoliday, data = ken_daily_orders)

model_6 <- lm(log_orders ~ day_of_week + is_PublicHoliday, data = ken_daily_orders)
model_7 <- lm(log_orders ~ day_of_week*is_PublicHoliday, data = ken_daily_orders)
model_8 <- lm(log_orders ~ day_of_week*is_PublicHoliday + is_SchoolHoliday, data = ken_daily_orders)

model_9 <- lm(log_orders ~ day_of_week * (is_PublicHoliday + is_SchoolHoliday), data = ken_daily_orders)


summary(model_3)
summary(model_4)
summary(model_5)
summary(model_6)

anova(model_1, model_3, model_4, model_5, model_9)
anova(model_1, model_6, model_7, model_8, model_9)

plot(model_9)
```

```{r visualizing public and schoolholidays}
ken_weekly_orders <- ken_daily_orders %>%
  mutate(week = floor_date(datum_creatie, unit = "week", week_start = 1)) %>%
  group_by(week) %>%
  summarise(
    weekly_order_count = sum(order_count),
    is_PublicHoliday = sum(is_PublicHoliday) > 0,
    is_SchoolHoliday = sum(is_SchoolHoliday) > 0
  ) %>%
  ungroup() %>%
  filter(week < max(week) & week > min(week)) # Exclude last and first weeks, as it is incomplete

# Retrieving holiday labels
public_holiday_labels <- public_holidays_nl %>%
  mutate(
    week = floor_date(date, unit = "week", week_start = 1),
    holiday_label = format(date, "%a")  # just the weekday abbreviation
  ) %>%
  group_by(week) %>%
  summarise(holiday_label = paste(unique(holiday_label), collapse = ", "))

# Get unique week-holiday combinations
school_weekly_holidays <- school_holidays %>%
  filter(date <= max(ken_daily_orders$datum_creatie)) %>%  # exclude any holidays after data ends
  mutate(week = floor_date(date, unit = "week", week_start = 1),
         holiday_name = gsub("Holidays", "", holiday_name),           # remove 'Holidays'
    holiday_name = trimws(holiday_name)) %>%
  distinct(week, holiday_name) %>%
  arrange(holiday_name, week)

school_weekly_holidays <- school_weekly_holidays %>%
  group_by(holiday_name) %>%
  arrange(week) %>%
  mutate(
    gap = as.numeric(week - lag(week, default = first(week))) > 7,
    group_id = cumsum(gap)
  ) %>%
  ungroup()

holiday_blocks <- school_weekly_holidays %>%
  group_by(holiday_name, group_id) %>%
  summarise(label_week = week[ceiling(n() / 2)], .groups = "drop") %>%
  rename(holiday_name = holiday_name)

# Adding labels to dataset
ken_weekly_orders_labeled <- ken_weekly_orders %>%
  left_join(public_holiday_labels, by = "week") %>% 
  left_join(school_holiday_labels, by = "week")
  
ggplot() +
  # School holiday shading
  geom_rect(data = subset(ken_weekly_orders, is_SchoolHoliday),
            aes(xmin = week - 3.5, xmax = week + 3.5, ymin = -Inf, ymax = Inf),
            fill = "#FF7F50", alpha = 0.2) +

  # Weekly trend line
  geom_line(data = ken_weekly_orders,
            aes(x = week, y = weekly_order_count),
            color = "#60A5FA", size = 1) +

  # Red crosses for public holidays
  geom_point(data = subset(ken_weekly_orders, is_PublicHoliday),
             aes(x = week, y = weekly_order_count),
             shape = 4, color = "red", size = 3) +
  
  # Label of Public holiday
  geom_text(data = subset(ken_weekly_orders_labeled, is_PublicHoliday),
          aes(x = week, y = weekly_order_count + 500, label = holiday_label),
          angle = 90, hjust = 0.5, vjust = 0, size = 2, color = "black") +

  # Single vertical label per school holiday block
  geom_text(data = holiday_blocks,
            aes(x = label_week, y = Inf, label = holiday_name),
            angle = 90, vjust = 0.7, hjust = 1, size = 2.5, color = "#F87171") +

  labs(title = "Weekly Order Volume with Holiday Periods",
       subtitle = "Orange background = School holidays; Red crosses = Public holidays",
       x = "Week", y = "Weekly Order Count") +
  theme_minimal()

ggsave(here("figures", "weekly_order_volume_and_holiday_periods.png"),width = 10, height = 6, dpi = 300)
```

```{r}
library(dplyr)
library(lubridate)

# Create weekly order data
ken_weekly_orders <- ken_daily_orders %>%
  mutate(week = floor_date(datum_creatie, unit = "week", week_start = 1)) %>%
  group_by(week) %>%
  summarise(
    weekly_order_count = sum(order_count),
    .groups = "drop"
  )

# Prepare school holiday weeks + names
school_weekly_holidays <- school_holidays %>%
  mutate(week = floor_date(date, unit = "week", week_start = 1)) %>%
  distinct(week, holiday_name) %>%
  arrange(holiday_name, week)

# Group contiguous holiday weeks (assign group ID)
school_weekly_holidays <- school_weekly_holidays %>%
  group_by(holiday_name) %>%
  arrange(week) %>%
  mutate(
    gap = as.numeric(week - lag(week, default = first(week))) > 7,
    group_id = cumsum(gap)
  ) %>%
  ungroup()

# Merge with weekly orders
ken_weekly_orders_labeled <- ken_weekly_orders %>%
  left_join(school_weekly_holidays, by = "week")

# Compute average volume for each holiday block
holiday_blocks <- ken_weekly_orders_labeled %>%
  filter(!is.na(holiday_name)) %>%
  group_by(holiday_name, group_id) %>%
  summarise(
    avg_holiday_volume = mean(weekly_order_count, na.rm = TRUE),
    first_week = min(week),
    .groups = "drop"
  )

# Get pre-holiday week order volume
pre_holiday_orders <- ken_weekly_orders %>%
  mutate(next_week = week + 7) %>%
  select(next_week, weekly_order_count) %>%
  rename(first_week = next_week, pre_holiday_volume = weekly_order_count)

# Join holiday blocks with pre-holiday volumes
holiday_reduction <- holiday_blocks %>%
  left_join(pre_holiday_orders, by = "first_week") %>%
  mutate(
    pct_reduction = 100 * (pre_holiday_volume - avg_holiday_volume) / pre_holiday_volume
  )

# Summarize average % reduction per holiday type
average_reduction_per_holiday <- holiday_reduction %>%
  group_by(holiday_name) %>%
  summarise(
    avg_pct_reduction = mean(pct_reduction, na.rm = TRUE),
    n_periods = n(),
    .groups = "drop"
  )

# View result
print(average_reduction_per_holiday)
```

## Rebound days

This is day that follows after a holiday or a long weekend (not a regular weekend, that is just a monday). (We saw earlier that on weekend days, the demand is very low, naturally)

```{r}
ken_daily_orders <- ken_daily_orders %>%
  arrange(datum_creatie) %>%
  mutate(
    day_of_week = wday(datum_creatie, label = TRUE, abbr = TRUE, week_start = 1),
    is_weekday = day_of_week %in% c("Mon", "Tue", "Wed", "Thu", "Fri"),
    was_yesterday_school_holiday = lag(is_SchoolHoliday, 1),

    is_rebound_day = if_else(
      is_weekday &
      is_PublicHoliday == 0 &
      is_SchoolHoliday == 0 &
      was_yesterday_school_holiday == 1,
      1, 0
    )
  ) %>%
  select(-was_yesterday_school_holiday)

"Correlation"
cor(ken_daily_orders %>% select(log_orders, is_rebound_day), use = "complete.obs")

"Linear regression"
model_1 <- lm(log_orders ~ day_of_week, data = ken_daily_orders)
model_6 <- lm(log_orders ~ day_of_week + is_rebound_day, data = ken_daily_orders)

summary(model_1)
summary(model_6)
anova(model_1, model_6)
```

## Influenza Season

|           |              |              |          |
|-----------|--------------|--------------|----------|
| 2021/2022 | Week 8 2022  | Week 20 2022 | 13 weken |
| 2022/2023 | Week 50 2022 | Week 11 2023 | 14 weken |
| 2023/2024 | Week 3 2024  | Week 11 2024 | 9 weken  |
| 2024/2025 | Week 3 2025  | Week 11 2025 | 9 weken  |

```{r}
  "Adding influenza season dataset"
# Define influenza periods
influenza_periods <- data.frame(
  year = c(2022, 2023, 2024, 2025),
  start_date = as.Date(c("2022-02-21", "2022-12-12", "2024-01-15", "2025-01-13")),
  end_date   = as.Date(c("2022-05-22", "2023-03-19", "2024-03-17", "2025-03-16"))
)

# Create daily calendar
influenza_dates <- data.frame(
  date = seq(as.Date("2022-01-01"), as.Date("2025-04-24"), by = "day"),
  in_influenza = 0
)

# Flag all days within influenza periods
for (i in 1:nrow(influenza_periods)) {
  influenza_dates$in_influenza[
    influenza_dates$date >= influenza_periods$start_date[i] &
    influenza_dates$date <= influenza_periods$end_date[i]
  ] <- 1
}

"Adding variable to ken_daily_orders dataframe"
ken_daily_orders <- ken_daily_orders %>%
  left_join(influenza_dates, by = c("datum_creatie" = "date"))

"Visualizing the periods"
library(ggplot2)
ken_daily_orders$datum_creatie <- as.POSIXct(ken_daily_orders$datum_creatie)
influenza_periods$start_date <- as.POSIXct(influenza_periods$start_date)
influenza_periods$end_date <- as.POSIXct(influenza_periods$end_date)

ggplot(ken_daily_orders, aes(x = datum_creatie)) +
  # Add shaded region for influenza periods
  geom_rect(data = influenza_periods,
            aes(xmin = start_date, xmax = end_date, ymin = -Inf, ymax = Inf),
            fill = "#F87171", alpha = 0.7, inherit.aes = FALSE) +
  # Plot order counts
  geom_line(aes(y = order_count), color = "#3B82F6") +
  labs(title = "Order Count Over Time with Influenza Periods",
       x = "Date", y = "Order Count") +
  theme_minimal()

"Correlation"
cor(ken_daily_orders %>% select(order_count, in_influenza), use = "complete.obs")
cor(ken_daily_orders %>% select(log_orders, in_influenza), use = "complete.obs")

"Linear regression"
model_8 <- lm(formula = order_count ~ day_of_week + in_influenza, data = ken_daily_orders)
model_9 <- lm(formula = order_count ~ day_of_week *in_influenza, data = ken_daily_orders)

model_8 <- lm(formula = log_orders ~ day_of_week , data = ken_daily_orders)
model_9 <- lm(formula = log_orders ~ day_of_week + in_influenza, data = ken_daily_orders)

summary(model_8)
summary(model_9)
anova(model_8, model_9)

```

## Hay fever

```{r}
# Create in_pollen_season variable: mid-May to mid-July each year
ken_daily_orders <- ken_daily_orders %>%
  mutate(in_pollen_season = if_else(
    format(datum_creatie, "%m-%d") >= "05-15" & format(datum_creatie, "%m-%d") <= "07-15",
    1, 0
  ))

# linear regression
model_basic <- lm(log_orders ~ day_of_week, data = ken_daily_orders)
model_pollen <- lm(log_orders ~ day_of_week + in_pollen_season, data = ken_daily_orders)

summary(model_basic)
summary(model_pollen)
anova(model_basic, model_pollen)
```

## Temperature & Humidity

```{r}
"Importing weather data"
weather_data <- read_excel(here("Data", "External_Variables", "open-meteo-avg-temp-humid.xlsx"),
    col_types = c("date", "numeric", "numeric"))

# renaming variables
weather_data <- weather_data %>%
  rename(
    temperature = `temperature_2m_mean (°C)`,
    humidity = `relative_humidity_2m_mean (%)`)

"Adding weather variables to dataframe"
ken_daily_orders <- ken_daily_orders %>% 
  left_join(weather_data, by = c("datum_creatie" = "time"))

"Visualization"
ggplot(ken_daily_orders, aes(x = temperature, y = order_count)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Order Count vs. Temperature")

"Correlation matrix"
cor(ken_daily_orders %>% select(order_count, temperature, humidity), use = "complete.obs")

"Linear regression"
model_basic <- lm(log_orders ~ day_of_week , data = ken_daily_orders)
model_temperature <- lm(log_orders ~ day_of_week + temperature, data = ken_daily_orders)
model_humidity <- lm(log_orders ~ day_of_week + humidity, data = ken_daily_orders)


summary(model_temperature)
summary(model_humidity)

anova(model_basic, model_temperature)
anova(model_basic, model_humidity)
```

```{r}
weekly_data <- ken_daily_orders %>%
  mutate(log_order_count = log(order_count)) %>%
  group_by(week) %>%
  summarise(
    temperature = mean(temperature, na.rm = TRUE),
    humidity = mean(humidity, na.rm = TRUE),
    log_order_count = mean(log_order_count, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(week != min(week), week != max(week))  # Remove first/last week

# Reshape for plotting
weekly_long <- weekly_data %>%
  pivot_longer(cols = c(log_order_count, temperature, humidity),
               names_to = "Variable",
               values_to = "Value")

# Plot
ggplot(weekly_long, aes(x = week, y = Value)) +
  geom_line(color = "steelblue", size = 1) +
  facet_wrap(~ Variable, ncol = 1, scales = "free_y") +
  labs(
    title = "Weekly average log(order count), temperature, and humidity",
    x = "Week",
    y = "Value"
  ) +
  theme_minimal()


ggplot(weekly_data, aes(x = temperature, y = log_order_count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "log(order count) vs Temperature")


test <- lm(log_order_count ~ day_of_week + temperature + humidity, data = ken_daily_orders)
summary(test)
```

# Time Series Analysis

## Daily level

```{r}
"Visualizing daily orders over time"
# Decomposition
decomposed <- decompose(ken_ts)
plot(decomposed)

# Checking for stationarity
adf.test(ken_daily_orders$order_count)

# Plot ACF
acf(ken_daily_orders$order_count, lag.max = 30)
# Plot PACF
pacf(ken_daily_orders$order_count, lag.max = 30)

```

## Weekly level

```{r}
# Aggregate to weekly level
ken_weekly_orders <- ken_daily_orders %>%
  mutate(week = floor_date(datum_creatie, unit = "week")) %>%
  group_by(week) %>%
  summarise(weekly_order_count = sum(order_count), .groups = 'drop') %>%
  arrange(week) %>%
  ungroup() %>%
  filter(week < max(week) & week > min(week)) # Exclude last and first weeks, as it is complete

# Create ts object
ken_weekly_ts <- ts(
  ken_weekly_orders$weekly_order_count,
  start = c(year(min(ken_weekly_orders$week)), week(min(ken_weekly_orders$week))),
  frequency = 52)

# Get decomposition graph
stl_decomp <- stl(ken_weekly_ts, s.window = "periodic")
plot(stl_decomp)

# Extract the components
components <- stl_decomp$time.series

# Get standard deviation for each component
apply(components, 2, sd)

# Or get the proportion of variance explained
var_components <- apply(components, 2, var)
prop_var <- var_components / sum(var_components)
prop_var

# Downloading and saving image to /figures folder
png(
  filename = here("figures", "stl_order_weekly_data_decomposition.png"),
  width = 1000,
  height = 800
)
plot(stl_decomp)
dev.off()

# Plot ACF
acf(ken_weekly_orders$weekly_order_count, lag.max = 30)
# Plot PACF
pacf(ken_weekly_orders$weekly_order_count, lag.max = 30)
```

## Structural Break Analysis

There seems to have been a astructural break around 2023. We test this with the Chow Test. This test consideras a linear model split into samples at a predeterined break point.

In The Quandt Likelihood Ratio test, it tries to determine the breakpoint itself by copmuting the Chow test at all possible break points. The largest Chow test statistic is then chosen.

```{r}
"Finding break point, only using time trend"
# Create time variable
ken_daily_orders$time <- 1:nrow(ken_daily_orders)

# Fit full model
full_model <- lm(order_count ~ time, data = ken_daily_orders)

# Find the breaking points
breaks_model <- breakpoints(order_count ~ time, data = ken_daily_orders)
summary(breaks_model)
ken_daily_orders$datum_creatie[c(737)] # Date = 2024-01-07

# Perform Chow test with found breaking point 
sctest(full_model, type = "Chow", point = 737) 
# --> p-val = 0.4468, not significant
```

The found structural break, was not significant. This is probably because seasonalities were not take into account. now to take seasonality, which are: day of week, public holidays and school holidays into account:

```{r}
"Finding break point, also using other variables"
# Where are breakpoints
breaks_model <- breakpoints(order_count ~ time + day_of_week + is_PublicHoliday + is_SchoolHoliday, data = ken_daily_orders)

# Summary gives you the break date(s)
summary(breaks_model)
ken_daily_orders$datum_creatie[c(463)] # Date = 2023-04-08

full_model2 <- lm(order_count ~ time + day_of_week + is_PublicHoliday + is_SchoolHoliday,
                 data = ken_daily_orders)

# Perform Chow test with found breaking point 
sctest(full_model2, type = "Chow", point = 463) 
# --> p-val = 4.308e-13, very significant
```

```{r adding break variables}
break_point <- 463

ken_daily_orders <- ken_daily_orders %>%
  mutate(
    day_index = row_number(),
    trend_pre_break = if_else(day_index <= break_point, day_index, 0),
    trend_post_break = if_else(day_index > break_point, day_index - break_point, 0)
  )
```

# Dataset for analysis

```{r breakpoint}
#write.csv(ken_daily_orders, file = here::here("Data", "ken_daily_orders_variables.csv"), row.names = FALSE)
```
